---
title: "CharacterLM"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
  html_document: default
date: '`r Sys.Date()`'
vignette: |
  %\VignetteIndexEntry{Vignette Title} %\VignetteEngine{knitr::rmarkdown} %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message=FALSE}
Sys.setenv(RETICULATE_PYTHON = "/home/alizaidi/anaconda3/envs/cntk-py35/bin/python")
library(cntk)
library(magrittr)
library(Matrix)
```

## Goals and Overview

This example demonstrates how to build a neural character language model with CNTK using regular plaintext data.

A neural language model uses a recurrent neural network to predict words (or characters) with a richer context than traditional n-gram models allow. In this implementation, a character is run through an LSTM and the output is then put through a fully-connected layer to predict the next output character. The model can learn to be extremely expressive as the context is progressively built-up with each letter run through the RNN. For even more expressiveness, we allow a stack of LSTMs where the output of each layer is put through the next layer as its input.

This example is inspired by Andrej Karpathy's blog post, ["The Unreasonable Effectiveness of Recurrent Neural Networks"](http://karpathy.github.io/2015/05/21/rnn-effectiveness/) and his accompanying code at https://github.com/karpathy/char-rnn. This example allows you to achieve similar results to those displayed in Karpathy's blog, but with the packed-sequence training efficiency that CNTK allows.

```{r}
hidden_dim <- 256
num_layers <- 2
minibatch_size <- 100
```

```{r}
get_data <- function(p, minibatch_size, data, char_to_ix, vocab_dim) {
  xi <- c()
  yi <- c()
  
  # the character LM predicts the next character so get sequences offset by 1
  for (i in p : p + minibatch_size) {
    xi <- c(xi, char_to_ix[data[i]])
    yi <- c(yi, char_to_ix[data[i + 1]])
  }
  
  # produce one-hot vectors
  X <- sparseMatrix(1:length(xi), xi) * 1 # * 1 to make numeric
  Y <- sparseMatrix(1:length(yi), yi) * 1
  
  # return a list of matrices for each of X (features) and Y (labels)
  list(list(X), list(Y))
}
```

```{r}
sample <- function(root, ix_to_char, vocab_dim, char_to_ix, prime_text = '',
                   use_hardmax = TRUE, length = 100, temperature = 1) {
  apply_temp <- function(p) {
    p <- p^temperature
    p / sum(p)
  }
  sample_word <- function(p) {
    if (use_hardmax) {
      return(exp(p) / sum(exp(p)))
    }
    # normalize probabilities and then take weighted sample
    p <- exp(p) / sum(exp(p))
    p <- apply_temp(p)
    sample(1:length(vocab_dim))
  }
  
  plen <- 1
  prime <- -1
  
  # start sequence with first input
  x <- matrix(0, ncol = vocab_dim)
  if (prime_text != '') {
    plen <- length(prime_text)
    prime <- char_to_ix[prime_text[0]]
  } else {
    prime <- sample(1:vocab_dim)
  }
  
  x[prime] <- 1
  arguments <- list(list(x), list(TRUE))
  
  # setup a vector for the output characters and add the initial prime text
  output <- c(prime)
  
  # loop through prime text
  for (i in 1:plen) {
    p <- root %>% func_eval(arguments)
    
    # reset
    x <- matrix(0, ncol = vocab_dim)
    if (i < plen - 1) {
      idx <- char_to_ix[prime_text[i + 1]]
    } else {
      idx <- sample_word(p)
    }
    
    output <- c(output, idx)
    x[idx] <- 1
    arguments = list(list(x), list(TRUE))
  }
  
  # loop through length of generated text, sampling along the way
  for (i in 1:length-plen) {
    p <- root %>% eval(arguments)
    idx <- sample_word(p)
    output <- c(output, idx)
    
    x <- matrix(0, ncol = vocab_dim)
    x[idx] <- 1
    arguments <- list(list(x), list(FALSE))
  }
  
  # convert numeric representation back to characters
  chars <- c()
  for (char in output) {
    chars <- c(chars, ix_to_char[toString(char)])
  }
  
  paste(chars, collapse = '')
}
```

```{r}
load_data_and_vocab <- function(path) {
  # load data
  data <- readChar(path, file.info(path)$size)[[1]]
  chars <- unique(data)
  data_size <- length(data)
  vocab_size <- length(chars)
  sprintf("data has %d characters, %d unique", data_size, vocab_size)
  
  char_to_ix <- list()
  ix_to_char <- list()
  for (i in 1:length(chars)) {
    char_to_ix[[ chars[i] ]] <- i
    ix_to_char[[ toString(i) ]] <- chars[i]
  }
  
  # write vocab for future use
  write(chars, paste(path, ".vocab", sep = ''))
  
  list(data, char_to_ix, ix_to_char, data_size, vocab_size)
}
```


```{r}
create_model <- function(output_dim) {
  Sequential(
    For(1:num_layers, function() {c(
      Sequential(Stabilizer(), Recurrence(LSTM(hidden_dim), go_backwards = FALSE))
    )}),
    Dense(output_dim)
  )
}
```

```{r}
create_inputs <- function(vocab_dim) {
  input_seq_axis <- CNTKAxis('inputAxis')
  input_sequence <- seq_input_variable(shape = vocab_dim, sequence_axis = input_seq_axis, name = 'input')
  label_sequence <- seq_input_variable(shape = vocab_dim, sequence_axis = input_seq_axis, name = 'label')
  
  list(input_sequence, label_sequence)
}
```

```{r}
train_lm <- function(training_file, epochs, max_num_minibatches) {
  # load data and vocab
  l <- load_data_and_vocab(training_file)
  data <- l[1]
  char_to_ix <- l[2]
  data_size <- l[3]
  vocab_dim <- l[4]
  
  # model the source targets inputs to the model
  sequences <- create_inputs(vocab_dim)
  input_sequence <- sequences[1]
  label_sequence <- sequences[2]
  
  # create the model and apply to input sequence
  model <- create_model(vocab_dim)
  z <- model(input_sequence)
  
  # setup criteria
  loss <- loss_cross_entropy_with_softmax(z, label_sequence)
  error <- classification_error(z, label_sequence)
  
  # instantiate trainer object
  lr_per_sample <- learning_rate_schedule(0.001, UnitType('sample'))
  momentum_time_constant <- momentum_as_time_constant_schedule(1100)
  learner <- learner_momentum_sgd(z$parameters, lr_per_sample, momentum_time_constant,
                                  gradient_clipping_threshold_per_sample = 5,
                                  gradient_clipping_with_truncation = TRUE)
  progress_printer <- ProgressPrinter(freq = 100, tag = 'Training')
  trainer <- Trainer(z, c(loss, error), learner, progress_printer)
  
  sample_freq <- 1000
  minibatches_per_epoch <- min(floor(data_size / minibatch_size),
                               floor(max_num_minibatches / epochs))
  
  # print out some useful training information
  log_number_of_parameters(z)
  sprintf("Running %d epochs with %d minibatches per epoch\n", epochs, minibatches_per_epoch)
  
  for (epoch in 1:epochs) {
    # Specify the mapping of input variables in the model to actual minibatch data to be trained with
    # If it's the start of the data, we specify that we are looking at a new sequence (True)
    mask = c(TRUE)
    for (batch in 1:minibatches_per_epoch) {
      minibatch <- get_data(batch, minibatch_size, data, char_to_ix, vocab_dim)
      arguments <- list(list('input' = minibatch[1], 'label' = minibatch[2]), mask)
      mask <- c(FALSE)
      trainer %>% train_minibatch(arguments)
      
      global_minibatch <- epoch * minibatches_per_epoch + batch
      if (global_minibatch %% sample_freq == 0) {
        print(sample(z, ix_to_char, vocab_dim, char_to_ix))
      }
    }
    
    model_filename <- paste("models/shakespeare_epoch", toString(epoch + 1), ".dnn", sep = "")
    func_save(model_filename)
    sprintf("Saved model to '%s'", model_filename)
  }
  
}
```

```{r}
load_and_sample <- function(model_filename, vocab_filename, prime_text = '', use_hardmax = FALSE,
                            length = 1000, temperature = 1.0) {
  model <- func_load(model_filename)
  
  # load vocab
  char_to_ix <- list()
  ix_to_char <- list()
  chars <- strsplit(readChar(vocab_filename, file.info(vocab_filename)$size))[[1]]
  for (i in 1:length(chars)) {
    char_to_ix[chars[i]] <- i
    ix_to_char[toString(i)] <- chars[i]
  }
  
  sample(model, ix_to_char, length(chars), char_to_ix, prime_text = prime_text, use_hardmax = use_hardmax,
         length = length, temperature = temperature)
}
```

```{r}
epochs <- 50
max_num_minibatches <- .Machine$integer.max
train_lm("../example-data/tinyshakespeare.txt", epochs, max_num_minibatches)

model_path <- paste("../models/shakespeare_epoch", toString(epochs), ".dnn", sep = "")
vocab_path <- "../example-data/tinyshakespeare.txt.vocab"

output <- load_and_sample(model_path, vocab_path, prime_text = 'T', use_hardmax = FALSE,
                          length = 100, temperature = 0.95)

write('output.txt', output)
```



















